\\Large The heat is on: Rising temperatures alter how and where mammals move

### 2025-08-30 writing/manuscript.Rmd
finished editing intro

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of hyperthermia [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016]. By altering mammals' movement speed and habitat use [jointly, their movement behavior: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behavior, such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. Behavioral changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more rapid and serious physiological damage [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, particularly if they have a large body size [@dyer_travel_2023; @verzuh_behavioural_2023]. Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. Consequently, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Uncertainty in future conditions and their consequences on ecosystems present a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behaviour and (2) the implications such changes and uncertainty will have on populations and species' adaptability and resiliency, as well as overall community structure. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019]. However, it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@ipcc_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].

<!-- why is understanding the present and predicting into the future necessary -->

Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to respond to change in a proactive manner [@sawyer_identifying_2009; @mccain_assessing_2019]. Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness. Predicting how climate change will affect how, when, and where animals will move is necessary for identifying landscape dynamics and value will evolve throughout the century. Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.

***tie to text above:***

***...*** Temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], altered phenology [@bunnell_global_2011; ], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and extreme weather [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022; @berger_climate_2018]. Changes in temperature thus affect how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

@@

This paper provides a framework for quantifying the effects of proximal air temperature on animal movement using quantitative methods that are flexible and statistically robust. We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and community levels.

>>>>

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of hyperthermia [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016]. By altering mammals' movement speed and habitat use [jointly, their movement behavior: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behavior [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. Behavioral changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more rapid and serious physiological damage [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, particularly if they have a large body size [@dyer_travel_2023; @verzuh_behavioural_2023]. Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with growing anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Preparing for and responding to future changes require a comprehensive understanding of how temperature will affect how and where mammals move, as well as the implications such changes will have on species' adaptability and overall community structure. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019]. However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behavior and quantitative predictions of how their behavior will change in the current century.

<!-- zoom in to western canada -->

Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022].

@@

This paper provides a quantifyies the effects of proximal air temperature on mammalian movement behavior, namely species' movement rates and habitat selection. We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)). Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature. We then pair the estimated responses with climate change projections to forecast the species' movement behavior during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017]. We discuss the consequences of these changes at the individual, population, and community levels. We also provide suggestions for future research, including sampling design to overcome some common data limitations.
